### 2025-02-26 manuscript/manuscript.qmd
references updated and conclusions revised. Some structural changes as well.

abstract: |
 This meta-analysis examines music emotion recognition (MER) models published between 2014 and 2024, focusing on predictions of valence, arousal, and categorical emotions. A total of 553 studies were identified, of which 96 full-text articles were assessed, resulting in a final review of 34 studies. These studies reported 204 models, including 86 for emotion classification and 204 for regression. Using the best-performing model from each study, we found that valence and arousal were predicted with reasonable accuracy (r = 0.67 and r = 0.81, respectively), while classification models achieved an accuracy of 0.xx. Across modeling approaches, linear and tree-based methods generally outperformed neural networks in regression tasks, whereas neural networks and support vector machines (SVMs) showed superior performance in classification tasks. We highlight key recommendations for future MER research, emphasizing the need for greater transparency, feature validation, and standardized reporting to improve comparability across studies.  
keywords          : "music, emotion, recognition, meta-analysis"
wordcount         : "5555"

@@

# A brief history of MER

@@

# The semantic gap in MER

The difficulty in predicting valence reflects a broader challenge in information retrieval. Specifically, relations between low-level predictors from music and text and the perceptual phenomena they model remain poorly understood, reaching a ceiling in prediction accuracy [@celma_foafing_2006]. To address this so-called *semantic gap*, researchers have attempted to identify new feature sets with greater relevance to emotion [@panda2020audio; @chowdhury2021perceived], combine low-, mid-, and high-level features using multimodal data [@celma_foafing_2006], or train neural networks to automatically learn features from audio [@zhang_bridge_2016]. Through these approaches, MER researchers attempt to shatter a so-called *glass ceiling* [@downie_music_2008] by establishing central predictors for emotions. To date, however, no study has systematically compared results of these diverse approaches.

@@

@@

@@

@@

Moving on the arousal, we carry out the same meta-analytical analysis applying the random-effects model to arousal. Table 3 describes the broad pattern of results in tabular format, and Figure 3 illustrates the spread and heterogeneity of all studies for arousal. The overall correlation across the studies using the best performing model out of each study (Max) is 0.809 (CI95% 0.740-0.860). If we examine all the models reported in each study, the correlation drops marginally, to 0.791 (CI95% 0.770-0.810), despite this analysis including about four times as many models as taking the best model out of each study. For arousal, even the baseline models seem to perform at a relatively high level. However, the indicators of heterogeneity are again high ($\tau^2$ = 0.141 and $I^2$=97.9%), which suggests that summary may be misleading. However, the analysis of asymmetry does not reveal significant issues (Eggers test, $\beta$ = 0.789 95%CI -4.87-6.45, _t_ = 0.273, _p_ = 0.788). If we remove the studies that are outside the 95%CI in heterogeneity, leaves this 13 studies in the summary that has _r_ = 0.826 (95%CI 0.806-0.845) with $\tau^2$ = 0.0042 and $I^2$ = 76.8%. In other words, no material difference to the results obtained with all 22 studies.

@@

The analysis of the subdivision of studies shows that there is no significant differences between the studies using different number of features (_Q_(2) = 5.20, _p_ = .074) despite the differing means (_r_ = 0.885 for studies with less than 30 features, _r_ = 0.735 for 30 to 300 features, and _r_ = 0.804 for studies utilising over 300 features). The differences in the techniques, however, does show statistically significant variance between subgroups (_Q_(3) = 10.83, _p_ = .0127). The Neural Nets (NN) achieve poor prediction of arousal (_r_ = 0.660) in comparison to other techniques. The caveat of this subgroup analysis is the small number of observations for four techniques. When other subgroupings were explored, such as the type of journal (Engineering with 13 studies vs Psychology with 9 studies), no significant differences were observed (_Q_(1) = 0.94, _p_ = .333).

@@

## Classification studies

We next evaluated classification studies. The number of emotion classes predicted varied between studies, ranging from two to eight. Across studies, most operationalizations corresponded to either affective dimensions of the circumplex model or discrete emotion labels mapping to those dimensions [e.g., happy, sad, nervous, and calm and similar variants used in @agarwal2021an; @alvarez2023ri; @yeh2014po]. Despite common adoption of the circumplex quadrants, studies differed in modeling dcisions: one study predicted quadrants in a multi-class problem [@panda2020no], whereas another predicted them in a series of binary classification tasks [@bhuvanakumar2023em]. Others predicted valence and arousal in separate tasks [@hu2022de; @zhang2016br], and some of the models analyzed only predicted arousal [e.g., all models in @zhang2017fe; the CART model in @zhang2016br]. Finally, some studies excluded specific quadrants [@hizlisoy2021mu] due to a lack of relevant stimuli, whereas others divided quadrants into multiple levels of valence and arousal [@nguyen2017an; @sorussa2020em].   Because only a few studies separately reported valence and arousal or quadrant classification, we averaged prediction success across them for each model to increase comaparability with other studies.

@@

Accounting for outliers among the classifications results removes 6 studie, affecting both performance ($MCC$ = 0.8941, 95%CI: 0.8282-0.9356) and heterogeneity scores ($\tau^2$ = 0.0569, $I^2$ = 97.7%). However, Eggers' test does not indicate an asymmetric funnel plot ($\beta$ = -19.769, 95%CI: -39.29--0.24). After aggregating across dimensions, some of the best models included only classified arousal (@zhang2016br; @zhang2017fe). To assess these studies' impact on overall performance, we conducted the same meta-analytic procedure after excluding them. Classification accuracy ($MCC$ = 0.8758, 95%CI: 0.7235-0.9468) does not change notably, nor does the rank order of model classes in terms of performance. Conseqeuntly, we report on all 12 in subsequent analyses.

@@

The number of features analyzed does not significantly impact results (_Q_(2) = 3.91, _p_ = 0.1419), though $MCC$ shows differences on averages based on the number of features (< 30 F: $MCC$ = 0.9293; 30-300 F: $MCC$ = 0.8458; >300 F: $MCC$ = 0.7754). Conversely, model classes differ significantly (_Q_(3) = 4.22, _p_ = 0.239). Specifiaclly, neural networks outperform all other model types ($MCC$ = 0.9307), followed by SVMs ($MCC$ = 0.870) and tree-based methods ($MCC$ = 0.854), then linear methods ($MCC$ = 0.728). Studies did not differ based on whether stimuli comprised one or multiple genres (_Q_(1) = 0.12, _p_ = 0.7324), or whether studies employed binary or multi-class models (_Q_(1) = 0.03, _p_ = 0.8694).

#### Notes to results: In the pre-registration, we promised the following:

@@

# Conclusion and Discussion

@@

These results offer several important insights. First, compared to a 2013 report by @barthet2013, the glass ceiling for valence has risen from _r_ = 0.51 to 0.67. Conversely, predictive accuracy for arousal has shown no improvement (reported already in 2013 at _r_ = 0.83). For classification, studies reached variable classification rates from 0.497 using _F~1~_ score [@sanden2011empirical] to 79% average precision [@trohidis2008multi] in the past, whereas now the accuracy is 0.930 (?).

<!-- These numbers above could be reinterpreted here, they are not comparable with our MCC -->

Second, there is no single modelling technique that seems to arise on top, although linear models and random forests perform best in regression studies, and neural networks and support vector machines excel in classification tasks. We note that the model accuracy is surprisingly little affected by the number of features or the dataset size, likely the result of cross-validation techniques used to avoid overfitting, but the heterogeneity of the materials used in different studies may also mask substantial differences. For instance, the smaller datasets tend to use linear models and deliver higher model accuracy than larger datasets and those with a large number of features. We surmise that this might relate to disciplinary difference, where the smaller datasets come from music psychology, which puts a premium on data quality (quality control of the ground-truth data and features) rather than on data size and modelling techniques. If quality control is succesful, it decreases the random error inherent in human judgements. Furthermore, the relevance and the precision of the features impacts the accuracy of the models. Unfortunately, in many cases, it is impossible to attribute the sources of error, as the studies so far have not compared the feature sets systematically, nor a comparison of the datasets with identical features has been carried out. In the future, it would be advantageous to systematically assess these sources of error in MER studies to allow us to focus on where significant improvements can be made.

More broadly, the present study revealed an uncomfortable large variation in overall quality control and reporting in MER studies. In many cases, the reporting was insufficient to determine the features or their sources used in the models, and a large number of studies were discarded due to insufficient information about the data, model architectures, or outcome measures.

<!-- recommendations 1: reporting -->

Future reports should contain viable information about the data, models, and success metrics. The modelling process should include a description of cross-validation, feature extraction technique, feature reduction (if used), and actual accuracy measures. We would recommend future studies to use Matthews correlation coefficient for classification studies as currently there are numerous metrics (_F~1~_, precision, recall, balanced accuracy, etc.), which do not operate in the same range.

<!-- feature validation? -->

Feature validation...? (Two extreme positions: a limited number of features that are perceptual validated (psychology) -- domain knowledge –– vs a large number of features where the models identify what is needed (deep learning/engineering).

<!-- recommendations 2: sharing -->

It would be beneficial to share models and features transparently. If a study relies on published datasets, as most do, sharing the actual model (as code/notebooks) and features would allow for more direct comparison of model techniques for the same dataset. Also, not all datasets are shared, or at least the sharing is only done partially. The copyright restrictions may limit the sharing of some datasets, but at least features and examples should be available.

<!-- size, type of data --> 
We also note that the majority of the datasets used in MER studies are relatively small, but comparable to datasets for modeling facial expressions (median N=502, @krumhuber2017) and speech expressions (median of 1287, @hashem2023). However, annotated datasets for visual scenes and objects tend to be much larger, often exceeding 100,000 annotated examples [@deng2009imagenet; @krishna2017visual]. Datasets of these magnitudes seem to be required for appropriate utilisation of deep learning algorithms, which may explain the modest results observed in the music emotion recognition studies.

The majority of the datasets are Western pop ... Music and annotators that represent the Global North do dominate the studies at the moment, and ... [@bornDiversifyingMIRKnowledge2020].

Are the emotion recognition targets valid from the psychological and application point of view? Prediction of valence and arousal is popular, but has limited usefulness in practical applications that aim to capture experiences with music that are more aligned with models of music-induced emotions (such as GEMS by @zentner2008emotions).

Future efforts to prioritize feature validation, standardized reporting, and transparent sharing of research materials are needed for consistent improvements in MER.

>>>>

abstract: |
 This meta-analysis examines music emotion recognition (MER) models published between 2014 and 2024, focusing on predictions of valence, arousal, and categorical emotions. A total of 553 studies were identified, of which 96 full-text articles were assessed, resulting in a final review of 34 studies. These studies reported 204 models, including 86 for emotion classification and 204 for regression. Using the best-performing model from each study, we found that valence and arousal were predicted with reasonable accuracy (r = 0.67 and r = 0.81, respectively), while classification models achieved an accuracy of 0.87 as measured with Matthews correlation coefficient. Across modeling approaches, linear and tree-based methods generally outperformed neural networks in regression tasks, whereas neural networks and support vector machines (SVMs) showed superior performance in classification tasks. We highlight key recommendations for future MER research, emphasizing the need for greater transparency, feature validation, and standardized reporting to improve comparability across studies.  
keywords          : "music, emotion, recognition, meta-analysis"
wordcount         : "5555"

@@

## A brief history of MER

@@

## The semantic gap in MER

The difficulty in predicting valence reflects a broader challenge in information retrieval. Specifically, relations between low-level predictors from music and text and the perceptual phenomena they model remain poorly understood, reaching a ceiling in prediction accuracy [@celma_foafing_2006]. To address this so-called *semantic gap*, researchers have attempted to identify new feature sets with greater relevance to emotion [@panda2020audio; @chowdhury2021perceived], combine low-, mid-, and high-level features using multimodal data [@celma_foafing_2006], or train neural networks to automatically learn features from audio [@zhang2016br]. Through these approaches, MER researchers attempt to shatter a so-called *glass ceiling* [@downie_music_2008] by establishing central predictors for emotions. To date, however, no study has systematically compared results of these diverse approaches.

@@

## Study identification

@@

#### Outlier analysis

@@

#### Reporting splits

@@

Moving on the arousal, we carry out the same meta-analytical analysis applying the random-effects model to arousal. Table 3 describes the broad pattern of results in tabular format, and Figure 3 illustrates the spread and heterogeneity of all studies for arousal. The overall correlation across the studies using the best performing model out of each study (Max) is 0.809 (CI95% 0.740-0.860). If we examine all the models reported in each study, the correlation drops marginally, to 0.791 (CI95% 0.770-0.810), despite this analysis including about four times as many models as taking the best model out of each study. For arousal, even the baseline models seem to perform at a relatively high level.

#### Outlier analysis

However, the indicators of heterogeneity are again high ($\tau^2$ = 0.141 and $I^2$=97.9%), which suggests that summary may be misleading. However, the analysis of asymmetry does not reveal significant issues (Eggers test, $\beta$ = 0.789 95%CI -4.87-6.45, _t_ = 0.273, _p_ = 0.788). If we remove the studies that are outside the 95%CI in heterogeneity, leaves this 13 studies in the summary that has _r_ = 0.826 (95%CI 0.806-0.845) with $\tau^2$ = 0.0042 and $I^2$ = 76.8%. In other words, no material difference to the results obtained with all 22 studies.

@@

#### Reporting splits

The analysis of the subdivision of studies shows that there is no significant differences between the studies using different number of features (_Q_(2) = 5.20, _p_ = .074) despite the differing means (_r_ = 0.885 for studies with less than 30 features, _r_ = 0.735 for 30 to 300 features, and _r_ = 0.804 for studies utilising over 300 features). The differences in the techniques, however, does show statistically significant variance between subgroups (_Q_(3) = 10.83, _p_ = .0127). The Neural Nets (NN) achieve poor prediction of arousal (_r_ = 0.660) in comparison to other techniques. The caveat of this subgroup analysis is the small number of observations for four techniques. When other sub-groupings were explored, such as the type of journal (Engineering with 13 studies vs Psychology with 9 studies), no significant differences were observed (_Q_(1) = 0.94, _p_ = .333).

@@

## Results for classification studies

We next evaluated classification studies. The number of emotion classes predicted varied between studies, ranging from two to eight. Across studies, most operationalizations corresponded to either affective dimensions of the circumplex model or discrete emotion labels mapping to those dimensions [e.g., happy, sad, nervous, and calm and similar variants used in @agarwal2021an; @alvarez2023ri; @yeh2014po]. Despite common adoption of the circumplex quadrants, studies differed in modeling decisions: one study predicted quadrants in a multi-class problem [@panda2020no], whereas another predicted them in a series of binary classification tasks [@bhuvanakumar2023em]. Others predicted valence and arousal in separate tasks [@hu2022de; @zhang2016br], and some of the models analyzed only predicted arousal [e.g., all models in @zhang2017fe; the CART model in @zhang2016br]. Finally, some studies excluded specific quadrants [@hizlisoy2021mu] due to a lack of relevant stimuli, whereas others divided quadrants into multiple levels of valence and arousal [@nguyen2017an; @sorussa2020em]. Because only a few studies separately reported valence and arousal or quadrant classification, we averaged prediction success across them for each model to increase comparability with other studies.

@@

Accounting for outliers among the classifications results removes 6 studies, affecting both performance ($MCC$ = 0.8941, 95%CI: 0.8282-0.9356) and heterogeneity scores ($\tau^2$ = 0.0569, $I^2$ = 97.7%). However, Eggers' test does not indicate an asymmetric funnel plot ($\beta$ = -19.769, 95%CI: -39.29--0.24). After aggregating across dimensions, some of the best models included only classified arousal (@zhang2016br; @zhang2017fe). To assess these studies' impact on overall performance, we conducted the same meta-analytic procedure after excluding them. Classification accuracy ($MCC$ = 0.8758, 95%CI: 0.7235-0.9468) does not change notably, nor does the rank order of model classes in terms of performance. Consequently, we report on all 12 in subsequent analyses.

@@

The number of features analyzed does not significantly impact results (_Q_(2) = 3.91, _p_ = 0.1419), though $MCC$ shows differences on averages based on the number of features (< 30 F: $MCC$ = 0.9293; 30-300 F: $MCC$ = 0.8458; >300 F: $MCC$ = 0.7754). Conversely, model classes differ significantly (_Q_(3) = 4.22, _p_ = 0.239). Specifically, neural networks outperform all other model types ($MCC$ = 0.9307), followed by SVMs ($MCC$ = 0.870) and tree-based methods ($MCC$ = 0.854), then linear methods ($MCC$ = 0.728). Studies did not differ based on whether stimuli comprised one or multiple genres (_Q_(1) = 0.12, _p_ = 0.7324), or whether studies employed binary or multi-class models (_Q_(1) = 0.03, _p_ = 0.8694).

Brief summary

<!--
Notes to results: In the pre-registration, we promised the following:

@@

-->

# Discussion and conclusions

@@

## Moderate but systematic increases in accuracy

The results of the meta-analysis offer several important insights. First, compared to a 2013 report by @barthet2013, the glass ceiling for valence has risen from _r_ = 0.51 to 0.67. Conversely, predictive accuracy for arousal has shown no improvement (reported already in 2013 at _r_ = 0.83). For classification, studies reached variable classification rates from 0.497 using _F~1~_ score [@sanden2011empirical] to 79% average precision [@trohidis2008multi] in the past, whereas now the accuracy is generally notably better, here _MCC_=0.87, albeit the differences in metrics.

<!-- Note. there is no clear conversion from Average precision to MCC, although we could estimate this from our data -->

Second, there is no single modelling technique that seems to arise on top, although linear models and random forests perform best in regression studies, and neural networks and support vector machines excel in classification tasks. We note that the model accuracy is surprisingly little affected by the number of features or the dataset size, likely the result of cross-validation techniques used to avoid overfitting, but the heterogeneity of the materials used in different studies may also mask substantial differences. For instance, the smaller datasets tend to use linear models and deliver higher model accuracy than larger datasets and those with a large number of features. We surmise that this might relate to disciplinary difference, where the smaller datasets come from music psychology, which puts a premium on data quality (quality control of the ground-truth data and features) rather than on data size and modelling techniques. If quality control is successful, it decreases the random error inherent in human judgements.

Third, there is little work on the relevance of the features and how these impact the accuracy of the models. Unfortunately, in many cases, it is impossible to attribute the sources of error to features (as opposed to target emotions obtained from participants), as the studies so far have not compared the feature sets systematically, nor a comparison of the datasets with identical features has been carried out. In the future, it would be advantageous to systematically assess these sources of error in MER studies to allow us to focus on where and how significant improvements can be made.

## Recommendations for future MER studies

Broadly speaking, the present study revealed an uncomfortably large variation in overall quality control and reporting in MER studies. In many cases, the reporting was insufficient to determine the features used in the models or their sources. Additionally, a significant number of studies had to be discarded due to a lack of information about the data, model architectures, or outcome measures. We summarize these issues below as recommendations for improving (1) reporting and transparency, (2) feature definition, (3) dataset scale and content, and (4) the selection of emotion frameworks.

<!-- ### Reporting and transparency -->

Future reports should contain viable information about the data, models, and success metrics. The modelling process should include a description of cross-validation, feature extraction technique, feature reduction (if used), and actual accuracy measures. We would recommend future studies to use Matthews correlation coefficient (_MCC_) for classification studies as currently there are numerous metrics (_F~1~_, precision, recall, balanced accuracy, etc.), which do not operate in the same range. The situation is regression studies tends to be less volatile, but model accuracy may be reported with error measures (Mean Squared Error, Root Mean Squared Error, or Mean Absolute Error, or MSE, RMSE, MAE), which cannot be converted to relative measure such as _R^2_ without knowing the variance of the data.

It would be beneficial to share models and features transparently. If a study relies on published datasets, as most do, sharing the actual model (as code/notebooks) and features would allow for a more direct comparison of model techniques for the same dataset. Also, not all datasets are shared [@akiki2021; @santana2020], or at least the sharing is only done partially. The copyright restrictions may limit the sharing of some datasets, but at least features and examples should be available.

<!-- ### Feature definition -->

One of the crucial aspects of modeling is the choice of features, their quantity, and their validation. In the present materials, we observe a continuum defined by two extremes: one approach relies on domain knowledge, starting with a limited number of features that are assumed to be perceptually relevant, while the other employs a large number of features, allowing the model to determine which are most relevant for prediction. The former is typically favored in music psychology studies, whereas the latter is more common in engineering and machine learning fields. Our results suggest that the domain knowledge-driven approach leads to the highest model accuracy across all studies and techniques. However, as our sample includes studies from the past 10 years, it is important to note that deep learning models have only gained prominence in this field since 2021. Consequently, it is too early to generalize that models based on domain knowledge, with a limited number of features and classical techniques, will continue to outperform machine learning approaches that utilize a large (300+) feature set. We believe that the relatively modest size of datasets available so far has prevented deep learning approaches from fully leveraging their potential to learn generalizable patterns in the data.

<!-- ### Dataset scale and content -->

We also note that the majority of the datasets used in MER studies are relatively small, but comparable to datasets for modeling facial expressions (median N=502, @krumhuber2017) and speech expressions (median of 1287, @hashem2023). However, annotated datasets for visual scenes and objects tend to be much larger, often exceeding 100,000 annotated examples [@deng2009imagenet; @krishna2017visual]. Datasets of these magnitudes seem to be required for appropriate utilisation of deep learning algorithms, which may explain the modest results observed in the music emotion recognition studies.

<!-- Western bias -->

The majority of the datasets are Western pop, which represent only a fraction of the musical styles consumed globally. Also, the annotators represent the Global North do dominate the studies at the moment with some exceptions [@gomez-canon2023; @zhangPMEmo2018]. This lack of diversity may contribute the success of the task but presents a significant limitation in our understanding of MER more broadly [@bornDiversifyingMIRKnowledge2020].

<!-- ### Emotion frameworks chosen -->

Finally, we note that the relevance of of the emotion frameworks used in MER is not always explicitly discussed. The majority of studies rely either on valence and arousal or some combination of basic emotion categories. However, the relevance of quadrants of the affective circumplex or basic emotions for music consumers can have limited usefulness in practical applications that aim to capture experiences with music that are more aligned with models of music-induced emotions such as GEMS by @zentner2008emotions or what emotions music are thought to express [@eerola2025what].

The present meta-analysis demonstrates that significant progress has been made toward developing accurate and scalable MER models over the past decade. Future efforts should prioritize feature validation, standardized reporting, the construction of larger and more diverse datasets, and the transparent sharing of research materials to ensure further consistent improvements in MER.
